\documentclass[11pt]{article}

\usepackage{amsmath,amsfonts,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{geometry}
\usepackage{enumitem}
\usepackage[table]{xcolor}
\usepackage{mdframed}
\usepackage{natbib}
\usepackage{array}
\usepackage{multirow}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{subcaption}

\geometry{margin=1in}

% ── Callout boxes ─────────────────────────────────────────────────────────────
\definecolor{shadeblue}{RGB}{235,243,251}
\definecolor{shadegray}{RGB}{245,245,245}
\definecolor{shadeyellow}{RGB}{255,252,230}

\mdfdefinestyle{keyresult}{
  backgroundcolor=shadeblue, linecolor=blue!40, linewidth=0.8pt,
  innerleftmargin=10pt, innerrightmargin=10pt,
  innertopmargin=8pt,  innerbottommargin=8pt,
  skipabove=8pt, skipbelow=8pt
}
\mdfdefinestyle{graybox}{
  backgroundcolor=shadegray, linecolor=gray!40, linewidth=0.8pt,
  innerleftmargin=10pt, innerrightmargin=10pt,
  innertopmargin=6pt,  innerbottommargin=6pt,
  skipabove=6pt, skipbelow=6pt
}

% ── Title ─────────────────────────────────────────────────────────────────────
\title{
  \textbf{FinStructBench:}\\[4pt]
  \large A Benchmark for Structured Information Retrieval\\
  from Financial Documents Using Graph-Verifiable Questions
}
\author{Agus Sudjianto\thanks{H2O.ai and Center for Trustworthy AI
  Through Model Risk Management, University of North Carolina Charlotte}
  \and Wingyan Lau\thanks{Knowlytix.ai}}
\date{}

\begin{document}
\maketitle

% =============================================================================
\begin{abstract}
We introduce \textbf{FinStructBench}, a benchmark for evaluating the ability of
large language models (LLMs) to extract, aggregate and reason over structured
information in financial documents.  Unlike existing QA benchmarks that rely on
human-annotated answers, FinStructBench generates questions \emph{automatically}
from a document's underlying knowledge graph, making ground truth
\emph{provably correct by construction}.  We define ten question categories of
increasing structural complexity---exact recall, threshold checking,
cross-reference, counting, contradiction detection, multi-hop reasoning,
absence detection, ranking, numeric computation and cross-table
aggregation---and provide five benchmark instances spanning major financial
report types: model validation (SR~11-7), fair lending (ECOA/HMDA), stress
testing (CCAR/DFAST), credit portfolio review (OCC) and Basel~III capital
adequacy (Pillar~3).  The benchmark supports two question modes:
\emph{deterministic} (template-based) and \emph{hybrid} (LLM-paraphrased for
linguistic diversity).  Across 454 questions evaluated in both modes,
the graph-based retrieval baseline achieves 100\% accuracy by design,
while Claude Sonnet~4 scores 58\% (deterministic) and 59\% (hybrid),
with per-category accuracy ranging from 94\% on absence detection to
30\% on contradiction.  A three-way ingestion comparison (regex-only,
LLM-only and hybrid) using Claude Opus~4 demonstrates that LLM-only
extraction recovers only 12.5\% of structured data, while hybrid
ingestion preserves 100\% of regex values and adds 22\% more entries
from prose---empirically validating the regex-first, LLM-augmented
architecture.  FinStructBench is released as an open-source toolkit
enabling researchers to generate benchmark instances from any
markdown-formatted financial document.
\end{abstract}

% =============================================================================
\section{Introduction}
\label{sec:intro}

Financial regulation demands precise information extraction from structured
documents: capital ratios must be reported to basis-point accuracy, adverse
impact ratios must be computed across protected classes and stress-test
projections must be traced across scenarios and time horizons.  Errors in
these tasks carry material consequences---regulatory findings, enforcement
actions and capital shortfalls.

Large language models are increasingly deployed for financial document
analysis, yet existing benchmarks for financial NLP focus predominantly on
sentiment analysis~\citep{malo2014good}, named entity recognition~\citep{alvarado2015domain}
or free-form question answering~\citep{chen2021finqa, zhu2021tat} where ground
truth is established by human annotation.  These benchmarks have three
limitations in the structured-retrieval setting:

\begin{enumerate}[leftmargin=2em]
  \item \textbf{Annotation cost and error.}  Human annotators make mistakes on
    precise numeric lookups, especially in dense tables with hundreds of cells.
  \item \textbf{Limited coverage.}  Hand-crafted questions test only the
    patterns the annotator anticipated, missing systematic failure modes.
  \item \textbf{No structural baseline.}  Without a provably correct retrieval
    system, it is impossible to distinguish LLM errors from ambiguous ground truth.
\end{enumerate}

We address all three limitations with \textbf{FinStructBench}, a benchmark
construction methodology that:
\begin{itemize}[leftmargin=2em]
  \item Ingests any markdown-formatted financial document into a
    \emph{DocumentGraph} comprising exact numeric memory (ENM) entries and
    knowledge graph (KG) triples;
  \item Automatically generates questions by mining graph topology, with
    ground truth defined by graph traversal---making it provably correct;
  \item Provides a \emph{graph-based retrieval baseline} (the graph itself)
    that achieves 100\% by construction, against which any LLM or agentic
    system can be compared.
\end{itemize}

Our initial release includes 454 questions across five financial report
types and ten complexity categories.  Results with Claude Sonnet~4 reveal a
41--42 percentage-point gap between graph-based retrieval (100\%) and LLM
performance (58--59\%), with per-category accuracy ranging from 94\% on
absence detection to 30\% on contradiction.  A three-way ingestion
comparison further demonstrates that LLM-only extraction recovers only
12.5\% of structured data, while a hybrid regex+LLM pipeline preserves
100\% of deterministic values and adds 22\% more entries from prose
extraction.  These findings demonstrate that current LLMs lack the
structured traversal capabilities required for reliable financial
document analysis and that deterministic ingestion is a quantitative
necessity, not merely a design preference.

% =============================================================================
\section{Related Work}
\label{sec:related}

\paragraph{Financial NLP Benchmarks.}
FinQA~\citep{chen2021finqa} and TAT-QA~\citep{zhu2021tat} evaluate numerical
reasoning over financial tables, but rely on human-annotated answers and focus
on arithmetic operations (addition, percentage change) rather than structural
graph operations.  ConvFinQA~\citep{chen2022convfinqa} extends to
conversational settings but inherits the same annotation limitations.
FLUE~\citep{shah2022flue} aggregates multiple financial NLP tasks but does not
address structured retrieval from regulatory documents.

\paragraph{Table QA and Table-Specialized Models.}
WikiTableQuestions~\citep{pasupat2015compositional} and
HybridQA~\citep{chen2020hybridqa} benchmark table understanding but use
general-domain tables without the precision requirements of financial
regulation.  SQA~\citep{iyyer2017search} addresses sequential table queries
but does not test cross-table reasoning.
TAPAS~\citep{herzig2020tapas} extends BERT to encode tables jointly with
questions, selecting cells and applying aggregation operators.
TaPEx~\citep{liu2022tapex} pre-trains a generative model as a neural SQL
executor, achieving strong results on WikiTableQuestions.
However, both are limited to single-table contexts and struggle with the
multi-step numerical reasoning required in financial analysis.

\paragraph{Document Understanding.}
DocVQA~\citep{mathew2021docvqa} and financial variants evaluate visual
document understanding, which is complementary to our text-based setting.
Layout-aware models such as LayoutLMv3~\citep{huang2022layoutlmv3}
jointly encode text, layout, and visual features, while
Donut~\citep{kim2022donut} performs OCR-free document understanding
end-to-end and Pix2Struct~\citep{lee2023pix2struct} learns to parse
screenshots into structured markup.  These models address the upstream
task of extracting clean tables from document images, which is
complementary to FinStructBench's focus on reasoning \emph{over}
already-extracted structured data.
Recent work on long-context evaluation~\citep{li2024long} tests retrieval
from long documents but uses synthetic ``needle-in-haystack'' probes rather
than domain-specific structural queries.

\paragraph{Structured Reasoning and Tool Use.}
Chain-of-Table~\citep{wang2024chainoftable} guides LLMs to iteratively
transform tables through atomic operations (select, group, sort) as
intermediate reasoning states, achieving strong results on table QA.
DATER~\citep{ye2023dater} decomposes both tables and questions into
simpler sub-problems for LLM reasoning.
StructGPT~\citep{jiang2023structgpt} constructs specialized interfaces
to collect evidence from structured data (tables, knowledge graphs,
databases) and lets LLMs focus on reasoning over the collected information.
ReAct~\citep{yao2023react} interleaves reasoning traces with tool-calling
actions, and Toolformer~\citep{schick2023toolformer} trains models to
self-select when to invoke external tools such as calculators and search
APIs.  These tool-augmented approaches are directly relevant to
FinStructBench's MCP server architecture, where the LLM calls
deterministic graph operations rather than extracting structured data
from raw text.

\paragraph{Graph-Augmented Retrieval.}
GraphRAG~\citep{edge2024graphrag} builds entity knowledge graphs from
source documents and pre-generates community summaries for query-focused
retrieval, outperforming vector-based RAG on global sensemaking tasks.
KGQA benchmarks~\citep{zhang2022subgraph} evaluate question answering over
pre-existing knowledge graphs.  FinStructBench differs from both
in that the graph is \emph{constructed from} the document using purely
deterministic rules, and the benchmark tests whether an LLM can
implicitly perform the same traversal from the raw text without access
to the graph.

\paragraph{Question Generation Methodologies.}
Automatic question generation (QG) from structured data spans a broad
spectrum of approaches.  At one extreme, classical benchmarks such as
FinQA~\citep{chen2021finqa}, TAT-QA~\citep{zhu2021tat}, and
FinanceBench~\citep{islam2023financebench} rely entirely on human expert
annotation---a process that is expensive (\$2+ per question), inherently
noisy, and difficult to scale.  At the other extreme, fully
LLM-driven approaches have gained traction: FinForge~\citep{matlin2026finforge}
uses a five-stage pipeline with Gemini~2.5 Flash to generate
semi-synthetic financial QA pairs, YourBench~\citep{shashidhar2025yourbench}
constructs zero-shot evaluation sets from arbitrary user documents, and
BenchAgents~\citep{butt2024benchagents} decomposes benchmark creation
into multi-agent planning, generation, verification, and evaluation stages.
While these LLM-based methods achieve high linguistic diversity, they
introduce non-determinism, contamination risk (the generating model may
have memorized benchmark data), and dependence on API availability.

Between these extremes, \emph{programmatic graph-topology mining}
generates questions by algorithmically traversing a knowledge graph
and rendering the resulting logical forms as natural language.
GrailQA~\citep{gu2021grailqa} enumerates SPARQL queries over Freebase by
traversing graph structure up to four relation hops, then crowdsources
natural-language paraphrases.
LC-QuAD~\citep{trivedi2017lcquad} follows a similar template-based SPARQL
generation pipeline over DBpedia.
In the financial domain,
KG-QAGen~\citep{tatarinov2025kgqagen} generates 20{,}139 QA pairs from
structured credit agreements by systematically varying complexity along
multi-hop retrieval depth, set operations, and answer plurality.
Dynamic-KGQA~\citep{dammu2025dynamickgqa} extracts semantically coherent
subgraphs from YAGO~4.5 and uses an LLM to generate questions, producing
fresh train/test splits each run to resist data contamination.
More recently, several benchmarks have formalized table-level reasoning
complexity: TableBench~\citep{wu2025tablebench} defines 18 sub-categories
of table reasoning,
TableEval~\citep{zhu2025tableeval} combines template-prompted and
role-prompted LLM strategies with clustering and human verification, and
GraphArena~\citep{tang2025grapharena} benchmarks LLMs on graph
computational problems (shortest path, connectivity) over diverse graph
topologies.

FinStructBench occupies a distinctive position within this landscape.
Like GrailQA and KG-QAGen, it mines graph topology programmatically
to generate questions with provably correct ground truth.  However, it
differs in three key respects: (i)~the knowledge graph is
\emph{constructed from} the document via deterministic ingestion rather
than being a pre-existing knowledge base; (ii)~it defines ten
specialized generator categories tailored to financial analyst workflows
rather than general SPARQL templates; and (iii)~no LLM or crowdsourcing
is involved at any stage of question creation, ensuring perfect
reproducibility, zero contamination risk, and zero marginal generation
cost.  The tradeoff is limited linguistic diversity in the generated
questions---a limitation we address with an optional hybrid mode that
uses LLM paraphrasing while preserving the deterministic ground-truth
guarantee (Section~\ref{sec:question_gen}).

% =============================================================================
\section{Benchmark Construction}
\label{sec:construction}

FinStructBench's construction pipeline has three stages: document ingestion,
question generation, and evaluation.  Each stage is fully automated and
domain-agnostic.  Critically, the entire pipeline is \emph{purely
deterministic}---it relies exclusively on regex pattern matching,
string parsing, and rule-based heuristics, with no machine learning or
LLM involved at any point.  This design is deliberate: because the
ground-truth graph is constructed by verifiable, transparent rules
rather than by a model that could hallucinate or make probabilistic
errors, the resulting answers are provably correct by construction.

\subsection{Document Ingestion}
\label{sec:ingestion}

Given a markdown-formatted financial document, the ingester produces a
\textbf{DocumentGraph} $\mathcal{G} = (\mathcal{E}, \mathcal{T}, \Phi)$
with three components:

\begin{description}[leftmargin=1.5em, labelindent=0em]
  \item[Exact Numeric Memory (ENM) $\mathcal{E}$:]  A key-value store
    mapping $(\textit{type}, \textit{id}) \to (\textit{value},
    \textit{hash})$ where each numeric value is stored with its SHA-256
    hash for integrity verification.  The \textit{type} is derived from
    the sanitized section heading (e.g.,
    \texttt{capital\_adequacy\_ratios}), and the \textit{id} is a
    composite key encoding the entity, any secondary label columns, and
    (for multi-column tables) the column name, separated by~\texttt{/}.
    For example, a stress test table with columns Scenario, Quarter,
    CET1, and Tier~1 produces keys such as
    (\texttt{capital\_projections},
    \texttt{Baseline/Q1\,2026/CET1}),
    ensuring unambiguous lookup even when the same metric name appears
    in multiple rows.  Each value is stored alongside its SHA-256 hash
    (computed over the IEEE-754 double representation) so that any
    silent corruption is detected at read time.

  \item[Knowledge Graph Triples $\mathcal{T}$:]  A set of
    $(\textit{head}, \textit{relation}, \textit{tail})$ triples extracted
    from table structure.  For each row the ingester emits:
    (i)~a \texttt{has\_\textit{col}} triple for every numeric cell,
    linking the entity to its string-formatted value;
    (ii)~a categorical triple for every non-entity label column;
    (iii)~a \texttt{passes} or \texttt{fails} triple if a boolean column
    is present; and
    (iv)~an \texttt{in\_section} triple linking the entity to its
    section tag.

  \item[Phase Encoders $\Phi$:]  Inequality functions that map a metric
    type to a comparison operation.  Each encoder stores the value range
    $[v_{\min}, v_{\max}]$ observed for that metric (used internally for
    geometric encoding) and exposes a \texttt{check\_inequality}$(v,
    \tau, \mathrm{op})$ method supporting $\geq$, $>$, $\leq$, and $<$.
    Encoders are inferred from column names using regex matching against a
    dictionary of 11 financial metric types (Table~\ref{tab:encoders}).
\end{description}

\begin{algorithm}[t]
\caption{Document Ingestion Pipeline}
\label{alg:ingest}
\begin{algorithmic}[1]
\REQUIRE Markdown document $D$
\ENSURE DocumentGraph $\mathcal{G} = (\mathcal{E}, \mathcal{T}, \Phi)$
\STATE Parse $D$ into sections $\{s_1, \ldots, s_k\}$ by \texttt{\#\#}/\texttt{\#\#\#} heading hierarchy
\STATE Infer phase encoders $\Phi$ from column names across all tables
\FOR{each section $s_i$}
  \FOR{each markdown table $T_j$ in $s_i$}
    \STATE Classify columns into \textit{entity}, \textit{numeric}, \textit{boolean}, \textit{label}
    \STATE Detect entity column by name heuristic; detect boolean column
    \FOR{each row $r$ in $T_j$}
      \STATE $e \gets$ entity value (or subsection name, or \texttt{row\_$i$})
      \STATE $e_{\text{full}} \gets e\,/\,\text{label}_1\,/\,\ldots\,/\,\text{label}_m$ \hfill\COMMENT{composite ID}
      \FOR{each numeric column $c$}
        \STATE $\textit{id} \gets e_{\text{full}}/c$ if multi-column, else $e_{\text{full}}$
        \STATE $\mathcal{E}[\text{section\_tag}, \textit{id}] \gets (v_c,\; \text{SHA-256}(v_c))$
        \STATE $\mathcal{T} \gets \mathcal{T} \cup \{(e, \texttt{has\_}c, v_c)\}$
      \ENDFOR
      \STATE Add label triples: $(e, \texttt{has\_}\ell, v_\ell)$ for each label column $\ell$
      \IF{boolean column value $\in$ \{true, yes, weak, fail\}}
        \STATE $\mathcal{T} \gets \mathcal{T} \cup \{(e_{\text{full}}, \texttt{fails}, \text{tag})\}$
      \ELSIF{$\in$ \{false, no, strong, pass\}}
        \STATE $\mathcal{T} \gets \mathcal{T} \cup \{(e_{\text{full}}, \texttt{passes}, \text{tag})\}$
      \ENDIF
    \ENDFOR
  \ENDFOR
  \STATE Extract bullet-list key-value pairs as metadata ENM entries and triples
\ENDFOR
\RETURN $\mathcal{G}$
\end{algorithmic}
\end{algorithm}

\paragraph{Step-by-Step Ground-Truth Construction.}
We now describe the complete process by which a markdown document is
transformed into a provably correct ground-truth graph, using a
concrete example.  Consider the following table appearing under the
heading \texttt{\#\# Capital Adequacy Ratios}:

\begin{mdframed}[style=graybox]\small\tt
| Metric~~~~~ | Value | Status |\\
|------------|-------|--------|\\
| CET1~~~~~~ | 12.5~ | Pass~~ |\\
| Tier\_1~~~~ | 14.3~ | Pass~~ |\\
| Total~~~~~ | 18.1~ | Fail~~ |
\end{mdframed}

\noindent
The ingester processes this table through five sequential steps:

\begin{enumerate}[leftmargin=2em]

\item \textbf{Section Parsing.}
  The document is split into hierarchical sections by \texttt{\#\#}
  and \texttt{\#\#\#} headings.  The heading ``Capital Adequacy Ratios''
  is sanitized into the section tag
  \texttt{capital\_adequacy\_ratios} by stripping leading numbering
  (e.g., ``3.1''), removing non-alphanumeric characters (except
  underscores and spaces), replacing spaces with underscores, and
  lowercasing.

\item \textbf{Phase Encoder Registration.}
  Before processing any tables, the ingester concatenates the full text
  of all sections and scans it against 11 regex patterns
  (Table~\ref{tab:encoders}).  Matching is performed on the entire
  section content---including entity names, column headers, and prose
  text---not just column names.  In this example, the entity names
  ``CET1'' and ``Tier\_1'' match the pattern
  \texttt{(?i)\textbackslash b(CET1|Tier.?1|capital.ratio|leverage)\textbackslash b},
  registering the \texttt{capital\_ratio}
  encoder with value range $[0, 30]$.  The encoder itself stores only
  the range $(v_{\min}, v_{\max})$; the threshold values used for
  question generation (e.g., $\geq 4.5$ for capital ratios) are
  defined separately in the Threshold question generator.  Encoders
  are registered once per metric type and shared across all sections.

\item \textbf{Column Classification.}
  Each column's values are analyzed to determine its type.
  In this example: ``Metric'' contains strings $\to$ \textit{label};
  ``Value'' contains numbers (100\% parseable as float) $\to$ \textit{numeric};
  ``Status'' contains exclusively pass/fail keywords $\to$ \textit{boolean}.
  The ``Metric'' column is further identified as the \textit{entity}
  column because it matches the name heuristic (a list of 16 common
  column names including ``Metric'').

\item \textbf{ENM Entry Construction.}
  For each row, the entity name and numeric value(s) are extracted.
  Since this table has a single numeric column, the ENM key is the
  entity name directly (no column suffix needed).  The ingester stores:
  \begin{align*}
    \mathcal{E}[\texttt{capital\_adequacy\_ratios},\; \texttt{CET1}]
      &= (12.5,\; \texttt{SHA-256}(12.5)) \\
    \mathcal{E}[\texttt{capital\_adequacy\_ratios},\; \texttt{Tier\_1}]
      &= (14.3,\; \texttt{SHA-256}(14.3)) \\
    \mathcal{E}[\texttt{capital\_adequacy\_ratios},\; \texttt{Total}]
      &= (18.1,\; \texttt{SHA-256}(18.1))
  \end{align*}
  Each SHA-256 hash is computed over the IEEE-754 double-precision
  representation and verified on every subsequent lookup, ensuring
  bit-exact integrity throughout the benchmark lifecycle.

  For tables with multiple numeric columns (e.g., columns AUC,
  Precision, Recall), the column name is appended to form a composite
  key: $(\texttt{type},\; \texttt{Entity/Column})$.  Tables with
  secondary label columns (e.g., Scenario and Quarter) produce
  keys such as $\texttt{Baseline/Q1\,2026/CET1}$.

\item \textbf{Triple Extraction.}
  The ingester emits knowledge graph triples that capture the full
  relational structure of each row.  Numeric and section triples use the
  base entity name as head; boolean triples use the full composite entity
  ID (which equals the base name when no secondary labels are present):
  \begin{itemize}[leftmargin=1.5em]
    \item \emph{Numeric triples:}
      For each numeric cell, a \texttt{has\_\textit{col}} triple is
      emitted using the base entity name:
      $(\texttt{CET1}, \texttt{has\_value}, \texttt{12.5})$,
      $(\texttt{Tier\_1}, \texttt{has\_value}, \texttt{14.3})$,
      $(\texttt{Total}, \texttt{has\_value}, \texttt{18.1})$.
    \item \emph{Boolean triples:}  The ``Status'' column is boolean.
      Values \texttt{pass}, \texttt{false}, \texttt{no}, and
      \texttt{strong} map to a \texttt{passes} relation paired with
      an \texttt{is\_strong} triple.  Values \texttt{fail},
      \texttt{true}, \texttt{yes}, and \texttt{weak} map to
      \texttt{fails} paired with \texttt{is\_weak}.  In this example:
      $(\texttt{CET1}, \texttt{passes}, \texttt{capital\_adequacy\_ratios\_CET1})$,
      $(\texttt{CET1}, \texttt{is\_strong}, \texttt{capital\_adequacy\_ratios})$;
      $(\texttt{Total}, \texttt{fails}, \texttt{capital\_adequacy\_ratios\_Total})$,
      $(\texttt{Total}, \texttt{is\_weak}, \texttt{capital\_adequacy\_ratios})$.
    \item \emph{Section triples:}
      $(\texttt{CET1}, \texttt{in\_section}, \texttt{capital\_adequacy\_ratios})$, etc.
  \end{itemize}
\end{enumerate}

\noindent
After all sections are processed, the resulting DocumentGraph is
self-contained: every numeric value in the source document has a
corresponding ENM entry (verifiable by text search), every
relational fact is encoded as a triple, and every applicable
threshold is registered as a phase encoder.  This completeness
guarantee is what makes the graph a \emph{provably correct} ground
truth---any question generated from this graph can be answered by
deterministic traversal, and the answer can be independently verified
against the source text.

\paragraph{Column Classification.}
The ingester classifies each column into one of four types using
value-pattern analysis:
\begin{itemize}[leftmargin=2em]
  \item \textbf{Numeric:} columns where $>$70\% of non-empty values parse
    as floats (after stripping \$, \%, commas, and parentheses for negative
    values).
  \item \textbf{Boolean:} columns where \emph{every} non-empty value is
    one of: pass, fail, yes, no, true, false, weak, strong
    (case-insensitive).
  \item \textbf{Label:} non-numeric, non-boolean string columns.
  \item \textbf{Entity:} the primary identifier column, selected by name
    heuristic from a list of 16 common column names (e.g., ``Feature'',
    ``Segment'', ``Scenario'', ``Product''), falling back to the first
    label column.
\end{itemize}

\noindent
This classification is conservative by design: mixed columns (e.g.,
\texttt{3.2~(Pass)}) are classified as label rather than boolean or
numeric, meaning the corresponding pass/fail relationship is not
captured in the graph.  Questions are generated only from data
that \emph{is} captured, so coverage may be incomplete but ground
truth is never wrong.

\paragraph{Phase Encoders.}
Phase encoders serve two roles: (i)~they define the value range for
normalizing numeric values to $[0, 1]$, and (ii)~they provide the
threshold values and comparison operators used by the Threshold
question generator.
Table~\ref{tab:encoders} lists the 11 encoder types, their
column-name patterns, and the default threshold used for question
generation (not the encoding range).

\begin{table}[t]
\centering
\caption{Phase encoder types: column-name patterns, default question
thresholds, and comparison direction.  Thresholds are domain-standard
regulatory or statistical bounds; the Direction column shows whether
values should be above ($\geq$) or below ($\leq$) the threshold.}
\label{tab:encoders}
\small
\begin{tabular}{llcl}
\toprule
Metric Type & Pattern & Threshold & Direction \\
\midrule
AUC & \texttt{auc} & 0.5 & $\geq$ \\
Accuracy & \texttt{acc, accuracy} & 0.5 & $\geq$ \\
AIR & \texttt{air} & 0.8 & $\geq$ \\
Ratio & \texttt{ratio} & 0.0 & $\geq$ \\
Rate & \texttt{rate, default.rate, approval.rate} & 0.05 & $<$ \\
Loss & \texttt{loss, logloss, brier} & 0.5 & $<$ \\
Capital Ratio & \texttt{cet1, tier1, capital.ratio, leverage} & 4.5 & $\geq$ \\
Importance & \texttt{importance, score, weight} & 0.1 & $\geq$ \\
Divergence & \texttt{divergence, distance, js, kl} & 0.3 & $\geq$ \\
P-value & \texttt{p.value, pvalue} & 0.05 & $<$ \\
Percentage & \texttt{pct, percent, \%} & 0.0 & $\geq$ \\
\bottomrule
\end{tabular}
\end{table}


\subsection{Question Generation}
\label{sec:question_gen}

Questions are generated automatically by mining the DocumentGraph topology.
Each generator implements a \texttt{generate(graph)} method that returns a
pool of candidate questions, each comprising: (i)~a natural-language
question string, (ii)~a deterministic graph-traversal function that
computes the ground-truth answer, (iii)~a structured LLM prompt with
format instructions, and (iv)~a typed answer format (float, int, bool,
set, or string).  We define ten categories of increasing structural
complexity.  Below we describe the precise mining algorithm for each
category, including the graph traversal patterns, filtering heuristics,
and sub-patterns that determine which questions are generated from a
given document.

\subsubsection{Category 1: Exact Recall}

\paragraph{Purpose.}  Test whether an LLM can look up a single numeric
value by its structured key---the most fundamental retrieval operation.

\paragraph{Mining algorithm.}
The generator iterates over every ENM entry
$(\textit{type}, \textit{id}) \to (\textit{value}, \textit{hash})$ in the
graph.  Each composite ENM~ID is parsed into a base entity name and an
optional column name by splitting on~\texttt{/}:
\texttt{Entity/Label1/.../Column} yields
$(\text{entity} = \texttt{Entity},\; \text{column} = \texttt{Column})$.
Entries whose entity name matches the synthetic pattern \texttt{row\_\textbackslash d+}
(generated when no entity column was detected) are skipped, since these
names do not appear in the source document and would produce unanswerable
questions.

For each surviving entry, the generator produces one question.  If a
column name is present, the question specifies both the entity and the
column (e.g., \emph{``What is the `AUC' value for `Model\_A' (category:
model\_validation)?''}).  If the ENM~ID has no column component (single
numeric column in the source table), the question asks for the entity's
value directly.

\paragraph{Graph answer function.}
$\mathcal{E}[\textit{type}, \textit{id}].\text{value}$ --- a single
key lookup with SHA-256 integrity verification.

\paragraph{Answer type.} \texttt{float}.  The LLM prompt instructs exact
reproduction of all decimal places.

\subsubsection{Category 2: Threshold}

\paragraph{Purpose.}  Test whether an LLM can locate a value and compare
it against a domain-specific regulatory or statistical bound.

\paragraph{Mining algorithm.}
For each ENM entry, the generator attempts to match it to one of 11 phase
encoder types using a three-stage process:

\begin{enumerate}[leftmargin=2em]
  \item \textbf{Word-boundary regex matching.}  The concatenation of the
    entry's category and entity~ID is tested against a list of compiled
    regex patterns with word boundaries (e.g., \texttt{\textbackslash bauc\textbackslash b},
    \texttt{\textbackslash bcet1\textbackslash b},
    \texttt{\textbackslash bcapital.ratio\textbackslash b}).  Word boundaries
    prevent false positives such as ``pair'' matching the AIR encoder.
  \item \textbf{Semantic exclusion.}  For the \texttt{capital\_ratio} encoder
    specifically, entities matching dollar-denominated field names
    (e.g., \texttt{dividend}, \texttt{provision}, \texttt{charge\_off},
    \texttt{rwa}, \texttt{goodwill}) are excluded even if they appear in a
    capital-related category, because these are dollar amounts rather than
    ratios.
  \item \textbf{Value plausibility.}  The entry's numeric value is checked
    against the encoder's threshold range.  Entries with
    $|\text{value}| > \max(\tau \times 100,\; 1000)$ (where $\tau$ is the
    largest threshold for that encoder) are rejected, filtering out dollar
    amounts that passed the regex match.
\end{enumerate}

\noindent
If all three checks pass and the matched encoder is registered in the
graph, the generator produces one boolean question for each threshold
defined in Table~\ref{tab:encoders} (e.g., AUC $\geq 0.5$,
AIR $\geq 0.80$, capital ratio $\geq 4.5$\%).

\paragraph{Graph answer function.}
$v \gets \mathcal{E}[\textit{type}, \textit{id}]$; then
$\phi_{\textit{encoder}}.\text{check\_inequality}(v, \tau, \text{op})$
$\to$ \texttt{bool}.

\paragraph{Answer type.} \texttt{bool}.

\subsubsection{Category 3: Cross-Reference}

\paragraph{Purpose.}  Test set intersection across two relation types,
requiring the LLM to identify entities appearing in two distinct
relational contexts.

\paragraph{Mining algorithm.}
The generator builds an inverted index mapping each relation~$r$ to its
set of head entities: $H_r = \{h : (h, r, \_) \in \mathcal{T}\}$.
High-cardinality, low-information relations (\texttt{in\_section},
\texttt{has\_value}, \texttt{has\_effect}) are excluded.  Relations with
fewer than 2 head entities are also excluded.

For each pair of surviving relations $(r_1, r_2)$, the generator computes
$H_{r_1} \cap H_{r_2}$.  Only pairs with $2 \leq |H_{r_1} \cap H_{r_2}|
\leq 50$ are retained; intersections below~2 are trivial, and those
above~50 are too noisy for meaningful evaluation.

\paragraph{Graph answer function.}
$\{h : (h, r_1, \_) \in \mathcal{T}\} \cap \{h : (h, r_2, \_) \in
\mathcal{T}\}$ $\to$ set of entity name strings.

\paragraph{Answer type.} \texttt{set\_str} (comma-separated list).

\subsubsection{Category 4: Counting}

\paragraph{Purpose.}  Test exhaustive enumeration and aggregation over
triple sets---a task where LLMs tend to undercount by stopping before
scanning all triples.

\paragraph{Mining algorithm.}
The generator produces four sub-patterns of counting questions:

\begin{enumerate}[leftmargin=2em, label=(\alph*)]
  \item \textbf{Count by relation.}  For each relation~$r$ (excluding
    noisy relations), count the total number of triples:
    $|\{(h,r,t) \in \mathcal{T}\}|$.  Relations with fewer than~2
    triples are skipped.

  \item \textbf{Count by (relation, tail).}  Group triples by
    $(r, t)$ pairs and count unique head entities per group:
    $|\{h : (h, r, t) \in \mathcal{T}\}|$.  Only groups with
    2--50 heads are retained.

  \item \textbf{Count unique heads.}  For each relation~$r$, count the
    number of distinct entities that participate:
    $|\{h : \exists\, t,\; (h, r, t) \in \mathcal{T}\}|$.  Requires
    $\geq 3$ unique heads.

  \item \textbf{Conditional count (above median).}  A multi-step pattern
    that combines triple queries with ENM aggregation.  For each
    $(r, t)$ group with $\geq 3$ head entities, and for each ENM column
    where at least 3 of those entities appear: (i)~compute the median of
    all values in that column (not just the overlapping entities);
    (ii)~count how many entities from the $(r, t)$ group have a column
    value strictly above the median.  Questions where the count equals
    zero or equals the full overlap size (trivial answers) are discarded.
\end{enumerate}

\paragraph{Graph answer function.}
Varies by sub-pattern: (a--c) use \texttt{query\_triples} with
pattern matching and \texttt{len()}; (d) chains
\texttt{query\_triples} $\to$ ENM iteration $\to$ \texttt{median}
$\to$ filtered count.

\paragraph{Answer type.} \texttt{int}.

\subsubsection{Category 5: Contradiction}

\paragraph{Purpose.}  Test global consistency checking---whether the LLM
can detect that a feature passes validation in some segments but fails in
others.

\paragraph{Mining algorithm.}
The generator calls \texttt{graph.find\_contradictions()}, which
identifies entities possessing both \texttt{passes} and \texttt{fails}
triples within the same feature group.  Feature groups are determined by
splitting the test-name string on~\texttt{\_} (first component) and then
on~\texttt{:} and~\texttt{/} (to extract the feature name from composite
test identifiers).  Only entities where the passing and failing tests
share the same group prefix are considered contradictions---this avoids
flagging entities that legitimately pass one test and fail a different,
unrelated test.

If contradictions exist, three sub-types of questions are generated:

\begin{enumerate}[leftmargin=2em, label=(\alph*)]
  \item \textbf{Count:} How many features have contradictory performance?
    (\texttt{int})
  \item \textbf{Set:} Which features have contradictions?
    (\texttt{set\_str})
  \item \textbf{Per-feature boolean:} For each feature that participates
    in any pass/fail triple (not just contradicting ones), ask whether it
    has contradictory performance.  This produces both true-expected and
    false-expected questions.  (\texttt{bool})
\end{enumerate}

\noindent
This category is only applicable to instances with boolean columns; in
our release, only the model validation instance generates contradiction
questions.

\paragraph{Graph answer function.}
Calls \texttt{find\_contradictions()} and extracts unique feature names
via prefix parsing.

\paragraph{Answer type.} \texttt{int}, \texttt{set\_str}, or \texttt{bool}.

\subsubsection{Category 6: Multi-Hop}

\paragraph{Purpose.}  Test chained multi-step reasoning that requires
maintaining intermediate results across sequential graph operations---the
most structurally complex category.

\paragraph{Mining algorithm.}
ENM entries are grouped by type, then by column within each type using
the \texttt{/}-separated structure of composite ENM~IDs.  Columns with
fewer than~3 non-aggregate entries are excluded (aggregate/summary rows
such as ``Total'', ``Grand Total'', ``Overall'' are detected by regex
and skipped to avoid trivially dominant extrema).  The generator produces
five sub-patterns:

\begin{enumerate}[leftmargin=2em, label=(\alph*)]
  \item \textbf{Argmax/min $\to$ cross-category lookup (\texttt{float}).}
    For each column~$c_1$ in type~$t_1$, find the entity with the
    highest (or lowest) value.  Then search for that same entity
    (matched by base name) in a different type~$t_2$, column~$c_2$.
    A question is generated only if the bridge entity exists in both
    types.  The answer is the numeric value from the second lookup.\\
    \emph{``Which entity has the highest `AUC' in `model validation'?
    What is that entity's `AIR' in `fair lending'?''}

  \item \textbf{Spread / range (\texttt{float}).}
    For each column within a single type, compute
    $\max(c) - \min(c)$ across all non-aggregate entities.  Columns
    where the spread is zero (all values identical) are skipped.\\
    \emph{``What is the range (highest minus lowest) of `AUC' values
    in `model validation'?''}

  \item \textbf{Argmax/min entity name (\texttt{str}).}
    For each column, return the name of the entity with the highest
    or lowest value.  Columns with tied extrema are skipped to ensure
    an unambiguous answer.\\
    \emph{``Which entity has the lowest `Loss\_Rate' in
    `credit portfolio'?''}

  \item \textbf{Conditional filter $\to$ count (\texttt{int}).}
    Chains a KG triple query with ENM threshold checking.  For each
    $(r, t)$ group with $\geq 3$ head entities, and for each ENM column
    where those entities appear: match the column to a phase encoder
    (using the same three-stage matching as Threshold), then count how
    many entities from the triple group meet the encoder's threshold.
    Requires at least~2 overlapping entities.\\
    \emph{``How many entities that `passes' `test\_X' also have
    `AUC' $\geq 0.5$ in `model validation'?''}

  \item \textbf{Argmax/min $\to$ cross-category threshold check
    (\texttt{bool}).}
    A three-step chain: (i)~find the entity with the highest/lowest
    value in column~$c_1$ of type~$t_1$ (skipping ties);
    (ii)~look up that entity's value for column~$c_2$ in type~$t_2$;
    (iii)~check whether the cross-looked-up value meets a phase encoder
    threshold.\\
    \emph{``Does the entity with the highest `AUC' in `model validation'
    meet the 4/5ths rule for `AIR' in `fair lending'?''}
\end{enumerate}

\paragraph{Graph answer function.}
Each sub-pattern iterates over the ENM, maintaining intermediate state
(the identity of the extremal entity, the cross-looked-up value) before
returning the final answer.

\paragraph{Answer type.} \texttt{float}, \texttt{str}, \texttt{int}, or
\texttt{bool} depending on sub-pattern.

\subsubsection{Category 7: Absence}

\paragraph{Purpose.}  Test hallucination resistance---whether the LLM can
correctly verify that an entity does \emph{not} appear in a given section,
rather than fabricating a plausible value.

\paragraph{Mining algorithm.}
The generator collects the set of base entity names per ENM type
(excluding synthetic \texttt{row\_N} and aggregate names).  It requires at
least two distinct types to generate questions.  Three sub-patterns of
boolean questions are produced:

\begin{enumerate}[leftmargin=2em, label=(\alph*)]
  \item \textbf{Cross-type absent (false-expected).}  For each target
    type~$t$, find entities that exist in some other type~$t'$ but not
    in~$t$.  These are plausible-sounding names (they exist somewhere in
    the document) that the LLM might hallucinate a value for in the
    wrong section.

  \item \textbf{Near-miss distractors (false-expected).}  For each
    entity present in type~$t$, generate string variants that do not
    exist:
    \begin{itemize}[leftmargin=1.5em]
      \item \emph{Prefix variants:}  Progressively truncate the entity
        name (e.g., \texttt{Segment\_A1} $\to$ \texttt{Segment\_A}).
        The prefix must be at least 3~characters long, must not itself
        be present in the target type, and must exist as a real entity
        in some other type (ensuring it is a plausible name, not random
        truncation).
      \item \emph{Suffix variants:}  Append one of three suffixes
        (\texttt{0}, \texttt{a}, \texttt{\_v2}) to a present entity
        name (e.g., \texttt{Model\_1} $\to$ \texttt{Model\_1a}).  The
        variant must not exist anywhere in the graph.
    \end{itemize}

  \item \textbf{Positive controls (true-expected).}  For each entity
    that \emph{is} present in type~$t$, generate a confirmation question.
    This balances the question pool to prevent a trivial ``always answer
    false'' strategy.
\end{enumerate}

\paragraph{Graph answer function.}
Scans all ENM keys of the target type, parsing each to its base entity
name, and returns \texttt{True} if a match is found, \texttt{False}
otherwise.

\paragraph{Answer type.} \texttt{bool}.

\subsubsection{Category 8: Ranking}

\paragraph{Purpose.}  Test whether the LLM can extract all values in a
column, sort them correctly, and identify the entity at a specific ordinal
position---stressing both exhaustive scanning and ordering accuracy.

\paragraph{Mining algorithm.}
ENM entries are grouped by $(\text{type}, \text{column})$.  Groups with
fewer than~3 non-aggregate entries or with any tied values are discarded
entirely (ties would make rank positions ambiguous).  Within each
surviving group, entries are sorted in descending order (rank~1 =
highest).  Questions are generated for a fixed set of key positions:

\begin{itemize}[leftmargin=2em]
  \item Rank 1 (highest) and rank $n$ (lowest), always included.
  \item Rank 2 (second highest), if $n \geq 3$.
  \item Median position ($\lfloor n/2 \rfloor + 1$), if $n \geq 5$.
\end{itemize}

\paragraph{Graph answer function.}
Collects all $(\text{entity}, \text{value})$ pairs for the target
$(\text{type}, \text{column})$, sorts descending, returns the entity name
at position~$k$.

\paragraph{Answer type.} \texttt{str} (entity name).

\subsubsection{Category 9: Numeric Computation}

\paragraph{Purpose.}  Test multi-step numeric reasoning: extracting two or
more values and performing arithmetic---a key failure mode identified in
recent work on table QA.

\paragraph{Mining algorithm.}
ENM entries are grouped by $(\text{type}, \text{column})$.  Two
sub-patterns are generated:

\begin{enumerate}[leftmargin=2em, label=(\alph*)]
  \item \textbf{Pairwise arithmetic (\texttt{float}).}
    For every pair of entities $(e_a, e_b)$ sharing the same column
    (generated via $\binom{n}{2}$ combinations), four operations are
    applied:
    \begin{itemize}[leftmargin=1.5em]
      \item \emph{Sum:} $v_a + v_b$
      \item \emph{Difference:} $v_a - v_b$
      \item \emph{Ratio:} $v_a / v_b$ (skipped if $|v_b| < 10^{-12}$)
      \item \emph{Percentage change:} $\frac{v_b - v_a}{|v_a|} \times 100$
        (skipped if $|v_a| < 10^{-12}$)
    \end{itemize}
    Requires $\geq 2$ entities in the column.

  \item \textbf{N-ary aggregate (\texttt{float}).}
    For each column with $\geq 3$ entities, compute:
    \begin{itemize}[leftmargin=1.5em]
      \item \emph{Mean:} arithmetic average of all values in the column.
      \item \emph{Median:} middle value (or average of two middle values).
    \end{itemize}
    The LLM prompt explicitly states the number of entities to extract,
    testing completeness of value extraction.
\end{enumerate}

\paragraph{Graph answer function.}
(a)~Two \texttt{lookup} calls followed by the arithmetic operation;
(b)~iteration over all ENM entries matching the $(\text{type},
\text{column})$ pair, then \texttt{mean} or \texttt{median}.

\paragraph{Answer type.} \texttt{float}.

\subsubsection{Category 10: Cross-Table Aggregation}

\paragraph{Purpose.}  Test whether the LLM can locate the same entity
across multiple table sections and perform arithmetic on the
corresponding values---a common task in financial analysis where metrics
are reported in different contexts (e.g., baseline vs.\ stress scenarios).

\paragraph{Mining algorithm.}
The generator builds an index mapping each base entity name to all its
ENM occurrences: $\text{entity} \to [(t_i, c_i, \text{eid}_i, v_i),
\ldots]$.  Entities appearing in fewer than~2 distinct ENM types are
skipped.  Two sub-patterns are generated:

\begin{enumerate}[leftmargin=2em, label=(\alph*)]
  \item \textbf{Cross-section sum (\texttt{float}).}
    For each column~$c$ that appears in multiple types for the same
    entity, sum the entity's values across all types:
    $\sum_i \mathcal{E}[t_i, \text{eid}_i]$ where each $\text{eid}_i$
    shares the same base entity and column.  Requires the column to
    appear in $\geq 2$ types.

  \item \textbf{Cross-section difference (\texttt{float}).}
    For each pair of types $(t_1, t_2)$ sharing a common column~$c$ for
    the same entity, compute
    $\mathcal{E}[t_1, e/c] - \mathcal{E}[t_2, e/c]$.
\end{enumerate}

\paragraph{Graph answer function.}
(a)~Multiple \texttt{lookup} calls (one per type), then summation;
(b)~two \texttt{lookup} calls, then subtraction.

\paragraph{Answer type.} \texttt{float}.

\paragraph{Sampling and Validation.}
Each generator produces a pool of all structurally valid candidates from
the graph---the pool sizes range from 8 (multi-hop in Basel Capital,
which has only 27 ENM entries) to 2{,}443 (cross-reference in Credit
Portfolio, which has 2{,}334 triples).  From each pool, up to $N$
questions are sampled uniformly without replacement using a fixed random
seed (default $N=10$, seed~42) for reproducibility.  Before inclusion,
every candidate is \emph{validated}: the graph answer function is
re-executed and its output compared to the stored ground truth.
Questions where the answer is \texttt{None}, an empty set, or
inconsistent with the stored ground truth are discarded.  This
double-check guards against bugs in the answer function and ensures
that the benchmark is self-consistent.

\paragraph{Ground Truth by Construction.}
A key property of FinStructBench is that ground truth is \emph{defined} by
graph traversal, not by human annotation.  The ingester's correctness
can be verified independently: we confirm that every stored ENM value
appears verbatim in the source markdown (1{,}572/1{,}572 entries verified
across all five instances).  Given correct ingestion, the answer function
is guaranteed correct by the determinism of key lookup, pattern matching,
and set operations.  This eliminates annotation noise and enables scaling
to arbitrarily large question sets.

\subsection{Design Rationale: Deterministic vs.\ Hybrid Generation}
\label{sec:design_rationale}

The question generation methodologies surveyed in
Section~\ref{sec:related} can be organized along two axes: the
\emph{source of structural logic} (how the question's answerable form is
determined) and the \emph{source of natural language} (how the question
is expressed).  Table~\ref{tab:qg_comparison} compares representative
approaches along five properties critical for benchmark integrity.

\begin{table}[t]
\centering
\caption{Comparison of question generation approaches.
\checkmark\ = fully satisfied, $\sim$\,= partially satisfied,
\texttimes\ = not satisfied.}
\label{tab:qg_comparison}
\small
\begin{tabular}{l@{\hskip 6pt}c@{\hskip 6pt}c@{\hskip 6pt}c@{\hskip 6pt}c@{\hskip 6pt}c}
\toprule
Approach & \rotatebox{60}{Provable GT} & \rotatebox{60}{Zero cost} &
  \rotatebox{60}{Reproducible} & \rotatebox{60}{No contam.} &
  \rotatebox{60}{Ling.\ diversity} \\
\midrule
Human annotation & \texttimes & \texttimes & \checkmark & \checkmark & \checkmark \\
LLM-driven & $\sim$ & \texttimes & \texttimes & \texttimes & \checkmark \\
Template + crowdsource & \checkmark & \texttimes & $\sim$ & \checkmark & \checkmark \\
Graph-topology (ours) & \checkmark & \checkmark & \checkmark & \checkmark & \texttimes \\
Hybrid (ours, optional) & \checkmark & \texttimes & $\sim$ & \checkmark & \checkmark \\
\bottomrule
\end{tabular}
\end{table}

FinStructBench's default \textbf{deterministic mode} prioritizes the first
four properties---provable ground truth, zero marginal cost, bit-exact
reproducibility, and zero contamination risk---at the expense of linguistic
diversity.  We justify this tradeoff on three grounds:

\begin{enumerate}[leftmargin=2em]
  \item \textbf{Ground-truth integrity.}  In financial regulation, an
    incorrect ground-truth label can lead to false confidence in a model's
    accuracy on safety-critical tasks.  Human annotation introduces noise
    (FinQA reports inter-annotator agreement of 92.2\%~\citep{chen2021finqa}),
    and LLM-generated labels require expensive verification.
    Deterministic graph traversal eliminates both sources of error.
  \item \textbf{Contamination resistance.}  LLM-generated benchmarks risk
    circular evaluation: the same model family that generated the questions
    may have memorized the patterns.  Dynamic-KGQA~\citep{dammu2025dynamickgqa}
    addresses this with fresh splits, but cannot guarantee that the
    generating model did not influence the question distribution.
    Our rule-based generators are immune to this concern.
  \item \textbf{Scalability.}  Adding a new benchmark instance requires
    only a markdown document---no API calls, no annotation budget, and no
    crowdsourcing pipeline.  The ten generators automatically produce a
    question pool whose size scales with the document's graph complexity.
\end{enumerate}

\paragraph{Optional Hybrid Mode.}
To address the linguistic diversity limitation, FinStructBench provides
an optional \textbf{hybrid mode} inspired by the GrailQA
pipeline~\citep{gu2021grailqa}: the deterministic generators produce the
question logic, graph answer function, and ground truth as before, and an
LLM is then used solely to paraphrase the template-generated natural
language into more diverse, realistic phrasing.  Crucially, the
paraphrase step modifies only the surface form of the question---the
ground truth, the graph answer function, and the LLM evaluation prompt
remain unchanged, preserving provable correctness.

The toolkit exposes this choice via a \texttt{question\_mode} parameter
(default: \texttt{"deterministic"}).  In hybrid mode, each question's
\texttt{natural\_language} field is replaced by an LLM-generated
paraphrase, while the \texttt{llm\_prompt} (the structured evaluation
prompt sent to the model under test) is left unchanged.  This design
ensures that:
\begin{itemize}[leftmargin=2em]
  \item The LLM under test sees a fixed, format-controlled prompt
    regardless of mode---results remain comparable.
  \item The paraphrase LLM is used only for benchmark
    \emph{presentation}, never for ground-truth \emph{computation}.
  \item Users can reproduce the deterministic baseline at any time by
    switching modes, with no change to scoring or ground truth.
\end{itemize}

\subsection{Evaluation Protocol}
\label{sec:eval_protocol}

Each question is evaluated against two systems:

\begin{enumerate}[leftmargin=2em]
  \item \textbf{Graph Baseline:}  The answer function is executed
    directly on the DocumentGraph via deterministic traversal (key
    lookups, triple pattern matching, set operations).  No learning or
    training is involved---the graph achieves 100\% by construction
    and serves as the upper bound.
  \item \textbf{LLM Under Test:}  The LLM receives the full markdown
    document wrapped in \texttt{<report>} tags as context, along with a
    system prompt instructing it to answer precisely using only the
    provided data and to copy numbers exactly as they appear.
    Each question includes a format instruction specifying the expected
    answer tag (e.g., \texttt{ANSWER: 13.232}).
\end{enumerate}

\paragraph{Response Parsing.}
LLM responses are parsed using type-specific parsers with two-stage
extraction:
\begin{itemize}[leftmargin=2em]
  \item \textbf{Float:}  First attempts regex on \texttt{ANSWER:\textbackslash s*<number>};
    falls back to the last float literal in the response.
  \item \textbf{Int:}  Same two-stage strategy; fallback extracts
    integers in the range $[1, 999]$.
  \item \textbf{Bool:}  Matches \texttt{ANSWER:\textbackslash s*(true|false|yes|no)};
    falls back to keyword detection (\texttt{passes}, \texttt{does not},
    \texttt{fails}).
  \item \textbf{Set:}  Splits the \texttt{ANSWER:} line on commas and
    strips quotes.
  \item \textbf{String:}  Returns the \texttt{ANSWER:} line verbatim.
\end{itemize}

\paragraph{Scoring.}
Scoring uses exact match with the following type-specific rules:
float answers require $|a - g| < 10^{-6}$;
integer answers require equality after casting;
boolean answers require equality;
set answers require exact set equality (superset answers are also
accepted);
string answers require case-insensitive equality.
Tuple-valued answers (multi-hop) compare element-wise with float
tolerance $10^{-3}$.

% =============================================================================
\section{Benchmark Instances}
\label{sec:instances}

We release five benchmark instances covering major financial document types
encountered in US banking regulation.  Each instance is a synthetic but
structurally realistic report: we first defined the table schemas, section
structure, and entity names to mirror actual regulatory filings, then
populated numeric cells with randomly generated values drawn from
domain-appropriate distributions (e.g., capital ratios from
$\mathcal{N}(12, 3)$ clipped to $[4, 25]$, approval rates from
$\text{Beta}(5, 2)$).  All instances are generated with fixed random
seeds for reproducibility.
Table~\ref{tab:instances} summarizes their characteristics.

\begin{table}[t]
\centering
\caption{FinStructBench instance statistics.}
\label{tab:instances}
\small
\begin{tabular}{lrrrcl}
\toprule
Instance & ENM & Triples & Questions & Categories & Regulatory Context \\
\midrule
Model Validation & 275 & 5{,}586 & 100 & 10 & SR 11-7 \\
Fair Lending & 1{,}139 & 2{,}014 & 90 & 9 & ECOA / HMDA \\
Stress Test & 476 & 1{,}417 & 90 & 9 & CCAR / DFAST \\
Credit Portfolio & 1{,}503 & 2{,}356 & 90 & 9 & OCC Guidelines \\
Basel Capital & 494 & 784 & 84 & 9 & Basel III Pillar 3 \\
\midrule
\textbf{Total} & \textbf{3{,}887} & \textbf{12{,}157} & \textbf{454} & --- & --- \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Model Validation (SR~11-7).}
A model risk management report evaluating a credit scoring model across
multiple segments.  Contains accuracy metrics (AUC, precision, recall, F1),
feature importance analysis (FANOVA), slicing performance by demographic
segments, and resolution analysis.  This is the only instance with
contradiction questions, as it contains per-feature pass/fail assessments
that can be inconsistent across segments.

\paragraph{Fair Lending (ECOA/HMDA).}
A fair lending analysis examining loan disposition patterns across protected
classes (race, gender, age).  Contains adverse impact ratios, pricing
analysis with statistical significance tests, geographic distribution
indicators, and model fairness metrics (demographic parity, equalized odds).
The largest instance by ENM entries (1{,}139), reflecting the high dimensionality
of cross-tabulated demographic data.

\paragraph{Stress Test (CCAR/DFAST).}
A capital adequacy stress test with macroeconomic scenarios (baseline,
adverse, severely adverse) projected across nine quarters.  Contains loan
loss projections by portfolio segment, pre-provision net revenue forecasts,
capital ratio trajectories, and capital action plans.

\paragraph{Credit Portfolio Review (OCC).}
A credit portfolio examination covering portfolio composition, credit quality
metrics, risk rating distributions, concentration analysis, allowance
adequacy, vintage analysis, migration matrices and stress loss projections.

\paragraph{Basel Capital Adequacy (Pillar~3).}
A Basel~III regulatory disclosure with capital composition, risk-weighted
assets by exposure class, capital ratios, credit risk parameters (PD, LGD,
EAD), market risk (VaR), operational risk (SMA) and leverage ratios.
The smallest instance by ENM entries (494) but with complex hierarchical
table structure.


% =============================================================================
\section{Experiments}
\label{sec:experiments}

\subsection{Setup}

We evaluate Claude Sonnet~4 (\texttt{claude-sonnet-4-20250514}) against
the graph baseline across all five instances and all ten question
categories.  The LLM receives the complete markdown document as context
(500--1{,}000 lines) along with a structured prompt for each question
specifying the expected answer format.  We use 10 questions per category
(where available), sampled with seed~42, yielding 454 questions per mode.

To assess the effect of question phrasing on LLM performance, we run
the full benchmark in two modes: \emph{deterministic} (template-generated
questions) and \emph{hybrid} (questions paraphrased by an LLM for
linguistic diversity while preserving ground truth; see
Section~\ref{sec:question_gen}).  The graph baseline achieves 100\%
accuracy in both modes by construction, confirming that paraphrasing
does not alter the ground truth.

\subsection{Main Results}

\begin{table}[t]
\centering
\caption{Overall benchmark results: Graph baseline vs.\ Claude Sonnet~4
in deterministic and hybrid question modes.}
\label{tab:main_results}
\small
\begin{tabular}{lccccc}
\toprule
Instance & Questions & Graph & Det.\ LLM & Hyb.\ LLM & $\Delta$ \\
\midrule
Model Validation & 100 & 100 (100\%) & 69 (69\%) & 71 (71\%) & $+2$ \\
Fair Lending & 90 & 90 (100\%) & 57 (63\%) & 59 (66\%) & $+2$ \\
Stress Test & 90 & 90 (100\%) & 27 (30\%) & 27 (30\%) & $0$ \\
Credit Portfolio & 90 & 90 (100\%) & 53 (59\%) & 57 (63\%) & $+4$ \\
Basel Capital & 84 & 84 (100\%) & 57 (68\%) & 54 (64\%) & $-3$ \\
\midrule
\textbf{Total} & \textbf{454} & \textbf{454 (100\%)} & \textbf{263 (58\%)} & \textbf{268 (59\%)} & $\mathbf{+5}$ \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:main_results} presents the overall results.  The graph
baseline achieves 100\% across all instances by construction.  Claude
Sonnet~4 achieves 58\% overall in deterministic mode and 59\% in
hybrid mode---a marginal improvement of 5~questions (+1~pp).
Per-instance scores range from 30\% (Stress Test) to 69--71\%
(Model Validation), reflecting substantial variation in document
complexity.

\subsection{Results by Question Category}

\begin{table}[t]
\centering
\caption{Results by question category (aggregated across all instances),
ordered by decreasing deterministic LLM accuracy.}
\label{tab:by_category}
\small
\begin{tabular}{lcccccc}
\toprule
Category & Graph Op. & Total & Graph & Det.\ LLM & Hyb.\ LLM & $\Delta$ \\
\midrule
Absence & $\nexists\,(h,r,t)$ & 50 & 50 & 47 (94\%) & 47 (94\%) & $0$ \\
Threshold & $\phi(v) \geq \tau$ & 50 & 50 & 37 (74\%) & 36 (72\%) & $-1$ \\
Multi-Hop & chain$(f_1, f_2)$ & 50 & 50 & 34 (68\%) & 33 (66\%) & $-1$ \\
Ranking & $\text{sort}(\{v_i\})[k]$ & 50 & 50 & 33 (66\%) & 33 (66\%) & $0$ \\
Cross-Reference & $S_1 \cap S_2$ & 50 & 50 & 29 (58\%) & 34 (68\%) & $+5$ \\
Exact Recall & $\mathcal{E}[k]$ & 50 & 50 & 26 (52\%) & 24 (48\%) & $-2$ \\
Cross-Table Agg. & $\sum_i f(T_i)$ & 44 & 44 & 21 (48\%) & 21 (48\%) & $0$ \\
Numeric Comp. & $f(v_1, v_2)$ & 50 & 50 & 17 (34\%) & 18 (36\%) & $+1$ \\
Counting & $|\{h : (h,r,t)\}|$ & 50 & 50 & 16 (32\%) & 19 (38\%) & $+3$ \\
Contradiction & $\exists\, \text{pass} \land \exists\, \text{fail}$ & 10 & 10 & 3 (30\%) & 3 (30\%) & $0$ \\
\midrule
\textbf{Total} & --- & \textbf{454} & \textbf{454} & \textbf{263 (58\%)} & \textbf{268 (59\%)} & $\mathbf{+5}$ \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:by_category} reveals a clear complexity gradient across
the ten question categories.  We highlight the key patterns:

\begin{itemize}[leftmargin=2em]
  \item \textbf{Absence (94\%):}  Verifying that an entity does not
    exist in the graph.  The LLM excels at this task because it
    requires confirming the non-existence of a specific value---a
    form of focused search that plays to the model's strengths.
  \item \textbf{Threshold (74\%):}  Simple inequality checks against a
    single value.  The LLM only needs to locate one number and compare
    it to a threshold.
  \item \textbf{Multi-Hop (68\%):}  Chained lookup requiring an
    intermediate computation (e.g., argmin/argmax) followed by a
    cross-type lookup.  Performance varies substantially across
    instances (30--90\%), suggesting sensitivity to table structure.
  \item \textbf{Ranking and Cross-Reference (58--66\%):}  Tasks
    requiring comparison across multiple entities or relation types.
    Cross-reference shows the largest hybrid improvement ($+5$),
    suggesting that more natural phrasing helps the LLM parse
    set-intersection queries.
  \item \textbf{Exact Recall (52\%):}  Precise numeric lookup.  Despite
    being a single-hop operation, LLMs frequently return nearby values
    from the same table or confuse similarly-named entities.
  \item \textbf{Cross-Table Aggregation (48\%):}  Aggregating values
    across multiple tables.  Requires identifying and combining
    information from different document sections.
  \item \textbf{Numeric Computation (34\%) and Counting (32\%):}
    Exhaustive enumeration or arithmetic over extracted values.  LLMs
    tend to undercount or introduce calculation errors.
  \item \textbf{Contradiction (30\%):}  Global consistency checking
    across segments---the most challenging category (though limited
    to 10~questions from the Model Validation instance).
\end{itemize}

\subsection{Per-Instance Breakdown}

\begin{table}[t]
\centering
\caption{Claude Sonnet~4 accuracy by instance and category
(deterministic mode).  Each cell shows correct/total.  Shading
indicates: {\colorbox{green!15}{$\geq$80\%}},
{\colorbox{yellow!30}{40--79\%}}, {\colorbox{red!15}{$<$40\%}}.}
\label{tab:heatmap}
\footnotesize
\setlength{\tabcolsep}{3pt}
\begin{tabular}{l*{10}{c}}
\toprule
& \rotatebox{70}{Absence} & \rotatebox{70}{Threshold} & \rotatebox{70}{Multi-Hop} & \rotatebox{70}{Ranking} & \rotatebox{70}{Cross-Ref} & \rotatebox{70}{Exact Recall} & \rotatebox{70}{Cross-Tbl Agg} & \rotatebox{70}{Numeric Comp} & \rotatebox{70}{Counting} & \rotatebox{70}{Contradiction} \\
\midrule
Model Valid. & \cellcolor{green!15}9/10 & \cellcolor{green!15}10/10 & \cellcolor{green!15}8/10 & \cellcolor{green!15}10/10 & \cellcolor{yellow!30}7/10 & \cellcolor{yellow!30}6/10 & \cellcolor{green!15}10/10 & \cellcolor{red!15}3/10 & \cellcolor{red!15}3/10 & \cellcolor{red!15}3/10 \\
Fair Lending & \cellcolor{green!15}10/10 & \cellcolor{green!15}10/10 & \cellcolor{yellow!30}6/10 & \cellcolor{yellow!30}7/10 & \cellcolor{yellow!30}5/10 & \cellcolor{yellow!30}6/10 & \cellcolor{yellow!30}7/10 & \cellcolor{red!15}2/10 & \cellcolor{yellow!30}4/10 & --- \\
Stress Test & \cellcolor{green!15}8/10 & \cellcolor{yellow!30}7/10 & \cellcolor{red!15}3/10 & \cellcolor{red!15}1/10 & \cellcolor{yellow!30}7/10 & \cellcolor{red!15}0/10 & \cellcolor{red!15}0/10 & \cellcolor{red!15}0/10 & \cellcolor{red!15}1/10 & --- \\
Credit Port. & \cellcolor{green!15}10/10 & \cellcolor{yellow!30}4/10 & \cellcolor{green!15}8/10 & \cellcolor{yellow!30}6/10 & \cellcolor{yellow!30}7/10 & \cellcolor{green!15}9/10 & \cellcolor{red!15}0/10 & \cellcolor{yellow!30}6/10 & \cellcolor{red!15}3/10 & --- \\
Basel Capital & \cellcolor{green!15}10/10 & \cellcolor{yellow!30}6/10 & \cellcolor{green!15}9/10 & \cellcolor{green!15}9/10 & \cellcolor{red!15}3/10 & \cellcolor{yellow!30}5/10 & \cellcolor{yellow!30}4/4 & \cellcolor{yellow!30}6/10 & \cellcolor{yellow!30}5/10 & --- \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:heatmap} shows the full instance $\times$ category
breakdown in deterministic mode.  Several patterns emerge:

\paragraph{Absence is consistently high.}
All five instances score $\geq$80\% on absence detection, confirming
that LLMs are reliable at verifying that a specific value does not
appear in the document.

\paragraph{Stress Test is consistently hard.}
Stress Test scores only 30\% overall, with 0/10 on exact recall,
numeric computation, and cross-table aggregation.  The dense
quarter~$\times$~scenario matrices cause systematic extraction failures.

\paragraph{Instance difficulty varies widely.}
Model Validation is the easiest instance (69\%), benefiting from
well-labeled tables and distinctive entity names.  Stress Test (30\%)
and Credit Portfolio (59\%) are harder due to wide matrices and
complex hierarchical structures.

\paragraph{Category difficulty is instance-dependent.}
Multi-hop ranges from 3/10 (Stress Test) to 9/10 (Basel Capital),
and cross-table aggregation ranges from 0/10 (Stress Test, Credit
Portfolio) to 10/10 (Model Validation), demonstrating that structural
complexity interacts with document layout.

\subsection{Deterministic vs.\ Hybrid Question Mode}
\label{sec:det_vs_hybrid}

\begin{table}[t]
\centering
\caption{Deterministic vs.\ hybrid LLM accuracy by category (aggregated
across all five instances, 454~questions per mode).  The graph baseline
achieves 100\% in both modes.}
\label{tab:det_vs_hybrid}
\small
\begin{tabular}{lccc}
\toprule
Category & Det.\ LLM & Hyb.\ LLM & $\Delta$ \\
\midrule
Absence & 47/50 (94\%) & 47/50 (94\%) & $0$ \\
Contradiction & 3/10 (30\%) & 3/10 (30\%) & $0$ \\
Counting & 16/50 (32\%) & 19/50 (38\%) & $+3$ \\
Cross-Reference & 29/50 (58\%) & 34/50 (68\%) & $+5$ \\
Cross-Table Agg. & 21/44 (48\%) & 21/44 (48\%) & $0$ \\
Exact Recall & 26/50 (52\%) & 24/50 (48\%) & $-2$ \\
Multi-Hop & 34/50 (68\%) & 33/50 (66\%) & $-1$ \\
Numeric Comp. & 17/50 (34\%) & 18/50 (36\%) & $+1$ \\
Ranking & 33/50 (66\%) & 33/50 (66\%) & $0$ \\
Threshold & 37/50 (74\%) & 36/50 (72\%) & $-1$ \\
\midrule
\textbf{Total} & \textbf{263/454 (58\%)} & \textbf{268/454 (59\%)} & $\mathbf{+5}$ \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:det_vs_hybrid} compares LLM performance across the two
question modes.  Hybrid paraphrasing yields a net gain of 5~questions
(+1~pp overall), but the effect is category-dependent:

\begin{itemize}[leftmargin=2em]
  \item \textbf{Cross-reference benefits most} ($+5$, from 58\% to
    68\%).  Template-generated cross-reference questions use rigid
    phrasing with nested quoting; the paraphrased versions use more
    natural set-intersection language that may better match the LLM's
    training distribution.
  \item \textbf{Counting improves modestly} ($+3$, from 32\% to 38\%).
    Natural phrasing of enumeration queries may reduce
    misinterpretation of the counting target.
  \item \textbf{Exact recall slightly degrades} ($-2$, from 52\% to
    48\%).  Paraphrasing may remove disambiguating cues (e.g., quoted
    column names) that help the LLM locate the correct cell.
  \item \textbf{Most categories are unaffected} (6 of 10 categories
    show $\leq$1 question difference), indicating that the question
    surface form has limited impact on LLM accuracy for most task types.
\end{itemize}

Table~\ref{tab:paraphrase_examples} illustrates the paraphrasing
effect: template artifacts (extra quoting, rigid phrasing) are replaced
with natural, professional language while preserving all entity names,
column names, and threshold values exactly.

\begin{table}[t]
\centering
\caption{Example questions in deterministic vs.\ hybrid mode.  The
hybrid mode produces more natural phrasing while preserving the exact
same meaning and all named entities.}
\label{tab:paraphrase_examples}
\small
\begin{tabular}{p{1.8cm}p{5.5cm}p{5.5cm}}
\toprule
Category & Deterministic (template) & Hybrid (paraphrased) \\
\midrule
Exact Recall &
  What is the \texttt{'Value'} value for \texttt{'Utilization'} (category: fanova\dots)? &
  What is the Value for Utilization in the fanova\_main\_effect\_curves category? \\[4pt]
Threshold &
  Does \texttt{'test/AUC'} (model\_accuracy) \textbf{meet} better than random (AUC $\geq$ 0.5)? &
  Does the \texttt{'test/AUC'} model\_accuracy \textbf{achieve} performance better than random chance\dots? \\[4pt]
Contradiction &
  Which features \textbf{have} contradictory performance across segments? &
  Which features \textbf{exhibit} contradictory performance when analyzed across different segments? \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Error Analysis}
\label{sec:errors}

We analyze the 191 LLM failures (deterministic mode) across
representative error modes:

\paragraph{Adjacent Cell Confusion.}
In exact recall failures, the LLM frequently returns a value from a
neighboring cell in the same table.  For example, when asked for
``Q1~2026/Non-Interest Income,'' Claude returns the Non-Interest Expense
value from the same row.  This error is especially prevalent in the
Stress Test instance (0/10 on exact recall), which uses wide matrices
with 9 quarterly columns.

\paragraph{Partial Enumeration.}
In the majority of counting failures (34 of 50 questions missed), the LLM
provides a count lower than the ground truth.  The model appears to
enumerate entities until it reaches a ``satisfying'' count rather than
exhaustively scanning all triples.

\paragraph{Cross-Table Blindness.}
Cross-table aggregation achieves 0/10 on both Stress Test and Credit
Portfolio, where the required values span multiple sections.  The LLM
typically finds values from one table but misses contributions from
other sections of the document.

\paragraph{Arithmetic Errors.}
Numeric computation failures (33 of 50 missed) frequently involve
correct value extraction but incorrect arithmetic---for example,
computing a ratio with the numerator and denominator swapped, or
rounding to the wrong precision.

\subsection{Ingestion Mode Comparison}
\label{sec:ingestion_modes}

The preceding experiments use a deterministic regex-based ingestion
pipeline to construct the ground-truth graph.  A natural question is
whether an LLM could replace or augment this pipeline.  We compare
three ingestion strategies across all five benchmark instances:

\begin{enumerate}[leftmargin=2em]
  \item \textbf{Regex} (default): Pure deterministic extraction using
    regex pattern matching, string parsing and value-pattern heuristics.
  \item \textbf{LLM-Only}: The LLM (Claude Opus~4) extracts all ENM
    entries and triples directly from each table and prose section,
    with no regex preprocessing.
  \item \textbf{Hybrid}: Regex extraction as the base, with LLM
    fallback for ambiguous column classification, entity resolution
    and prose relation extraction.
\end{enumerate}

\paragraph{Metrics.}
We measure: (i)~ENM and triple counts per mode, (ii)~pairwise Jaccard
index over ENM key sets and triple sets, (iii)~value match rate (exact
numeric agreement on shared ENM keys), (iv)~regex recovery rate
(fraction of regex ENM entries recovered by each alternative mode),
and (v)~numeric precision (exact match and $<10^{-6}$ tolerance on
shared keys).

\begin{table}[t]
\centering
\caption{ENM and unique triple counts by ingestion mode across five
instances.  Triple counts are deduplicated (cf.\ Table~\ref{tab:instances},
which reports raw triples including duplicates).  The LLM-only mode
recovers only 12.5\% of regex ENM entries, while hybrid recovers all
regex entries and adds prose-extracted data.}
\label{tab:ingestion_counts}
\small
\begin{tabular}{lrrrrrr}
\toprule
& \multicolumn{3}{c}{ENM Entries} & \multicolumn{3}{c}{Triples} \\
\cmidrule(lr){2-4} \cmidrule(lr){5-7}
Instance & Regex & LLM & Hybrid & Regex & LLM & Hybrid \\
\midrule
Model Validation   & 275   & 77  & 1{,}107 & 2{,}109 & 159   & 3{,}041 \\
Fair Lending       & 1{,}139 & 109 & 1{,}140 & 1{,}611 & 279 & 1{,}594 \\
Stress Test        & 476   & 35  & 480     & 1{,}243 & 105   & 1{,}305 \\
Credit Portfolio   & 1{,}503 & 150 & 1{,}520 & 2{,}256 & 422 & 2{,}340 \\
Basel Capital      & 494   & 116 & 498     & 778     & 286   & 803 \\
\midrule
\textbf{Total}     & \textbf{3{,}887} & \textbf{487} & \textbf{4{,}745}
                   & \textbf{7{,}997} & \textbf{1{,}251} & \textbf{9{,}083} \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[t]
\centering
\caption{Pairwise Jaccard index and recovery rates.  Regex$\leftrightarrow$LLM
overlap is near zero because the LLM uses different naming conventions.
Hybrid preserves all regex values exactly (100\% value match) while
recovering 97.2\% of regex ENM entries on average.}
\label{tab:ingestion_jaccard}
\small
\begin{tabular}{lcccccc}
\toprule
& \multicolumn{2}{c}{Regex$\leftrightarrow$LLM} & \multicolumn{2}{c}{Regex$\leftrightarrow$Hybrid} & \multicolumn{2}{c}{Recovery (\%)} \\
\cmidrule(lr){2-3} \cmidrule(lr){4-5} \cmidrule(lr){6-7}
Instance & ENM $J$ & Triple $J$ & ENM $J$ & Triple $J$ & LLM & Hybrid \\
\midrule
Model Validation   & 0.048 & 0.023 & 0.207 & 0.084 &  5.8 &  86.2 \\
Fair Lending       & 0.000 & 0.043 & 0.999 & 0.895 &  0.0 & 100.0 \\
Stress Test        & 0.008 & 0.003 & 0.992 & 0.952 &  0.8 & 100.0 \\
Credit Portfolio   & 0.000 & 0.012 & 0.989 & 0.793 &  0.0 & 100.0 \\
Basel Capital      & 0.020 & 0.057 & 0.992 & 0.813 &  2.4 & 100.0 \\
\midrule
\textbf{Mean}      & \textbf{0.015} & \textbf{0.028} & \textbf{0.836} & \textbf{0.707} & \textbf{1.8} & \textbf{97.2} \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{LLM-only extraction is catastrophically incomplete.}
Table~\ref{tab:ingestion_counts} shows that the LLM-only mode extracts
only 487 of 3{,}887 regex ENM entries (12.5\%) and 1{,}251 of 7{,}997
triples (15.6\%).  Table~\ref{tab:ingestion_jaccard} reveals that even
the entries the LLM does extract have near-zero overlap with the regex
graph: the mean ENM Jaccard index is 0.015 and the mean regex recovery
rate is just 1.8\%.  This near-total disjointness arises because the
LLM generates different entity identifiers (e.g., ``CET1\_Capital\_Ratio''
vs.\ the regex-produced ``cet1\_capital\_ratio/capital\_ratios''),
different section tags and tends to summarize rather than exhaustively
enumerate all table cells.

\paragraph{Hybrid mode is a strict superset of regex.}
The hybrid mode recovers 97.2\% of regex ENM entries on average
(100\% on four of five instances), with a 100\% value match rate---every
shared numeric value is bit-identical.  The hybrid graph is strictly
larger: 4{,}745 ENM entries (+22\% over regex) and 9{,}083 triples
(+14\%), reflecting the additional prose-extracted data contributed by
the LLM component.  The one outlier is Model Validation (86.2\%
recovery, ENM Jaccard 0.207), where the LLM's column classification
differs from the heuristic on several ambiguous tables, leading to
different entity-id construction for some entries.

\paragraph{Numeric precision is exact.}
On all shared ENM keys between regex and hybrid modes, 100\% of
numeric values match exactly (within floating-point representation).
The hybrid mode introduces zero numeric errors across all 3{,}887
shared entries, confirming that the SHA-256 integrity-checked storage
preserves values regardless of ingestion mode.

\paragraph{Implications.}
These results provide empirical justification for the regex-first
architecture: the deterministic pipeline is not merely a design
preference but a quantitative necessity.  An LLM-only approach loses
87\% of structured data and produces a graph with fundamentally
different structure.  The hybrid mode demonstrates that the optimal
architecture uses regex as the backbone for exhaustive, precise
extraction, with the LLM serving as an augmentation layer for
content that falls outside the regex pipeline's scope.

% =============================================================================
\section{Discussion}
\label{sec:discussion}

\paragraph{Implications for Financial AI.}
The 41--42 percentage-point gap between graph-based retrieval (100\%)
and LLM performance (58--59\%) has direct implications for the deployment
of AI in financial regulation.  While the LLM performs well on absence
detection (94\%) and threshold checking (74\%), accuracy drops sharply on
tasks requiring exhaustive enumeration, arithmetic or cross-table
reasoning (30--48\%).  These findings suggest that LLM-only approaches
are insufficient for reliable financial document analysis.  Hybrid
architectures that combine LLM understanding with structured graph
retrieval are likely necessary for production-grade deployment.

\paragraph{Why Rule-Based Ingestion?}
FinStructBench's ingestion pipeline is deliberately rule-based (regex
pattern matching, string parsing, value-pattern heuristics) rather
than using the SOTA extraction techniques surveyed in
Section~\ref{sec:related}.  The ingestion mode comparison in
Section~\ref{sec:ingestion_modes} provides direct empirical evidence:
an LLM-only pipeline recovers only 12.5\% of structured data with
near-zero overlap (Jaccard~$= 0.015$).  Beyond this empirical
finding, the choice is driven by three requirements specific to
benchmark construction:

\begin{enumerate}[leftmargin=2em]
  \item \textbf{Provable correctness.}  The ground truth must be
    verifiably correct by construction.  LLM-based extraction
    (e.g., GPT-4 or Claude applied to table parsing) achieves
    68--74\% on FinQA~\citep{chen2021finqa}---accurate enough for many
    applications, but insufficient as a ground-truth oracle.  Any
    extraction error would propagate into incorrect benchmark answers,
    undermining the benchmark's purpose.  Rule-based parsing achieves
    near-100\% precision on well-structured markdown tables because
    the format is unambiguous.

  \item \textbf{Determinism and reproducibility.}  Given the same
    document, the pipeline must produce the identical graph on every
    run.  LLM outputs are inherently stochastic (even at temperature~0,
    sampling artifacts can vary across API versions), making them
    unsuitable for a reproducible ground-truth engine.  Rule-based
    parsing is fully deterministic.

  \item \textbf{Transparency and auditability.}  In a regulated
    financial setting, every ground-truth answer must be traceable to
    a specific cell in the source document.  The SHA-256 integrity hash
    on each ENM entry provides a cryptographic audit trail from answer
    to source value.  LLM-based extraction is opaque---there is no
    equivalent mechanism to verify that an extracted value matches the
    source.
\end{enumerate}

\noindent
The tradeoff is generalization: the rule-based pipeline requires
well-structured markdown with pipe-delimited tables.  It does not
handle prose-embedded numbers, scanned PDFs or irregular layouts
that layout-aware models like LayoutLMv3~\citep{huang2022layoutlmv3}
or Donut~\citep{kim2022donut} can process.  For real-world
deployment beyond the benchmark, we provide a \textbf{hybrid
ingestion mode} (described below) that uses an LLM to assist with
column classification, entity resolution and relation extraction,
while preserving the deterministic ENM storage and integrity hashing.

\paragraph{Hybrid Ingestion Architecture.}
To bridge the gap between rule-based precision and LLM-based
flexibility, the toolkit supports a hybrid ingestion mode
(\texttt{-{}-ingest-mode hybrid}) in which the LLM augments
three specific steps of the pipeline:

\begin{itemize}[leftmargin=2em]
  \item \textbf{Column classification:}  When heuristic classification
    is ambiguous (e.g., mixed columns, domain-specific labels), the
    LLM is prompted to classify each column as entity, numeric,
    boolean or label, given the table headers and a sample of rows.
  \item \textbf{Entity resolution:}  When the entity column cannot be
    detected by name heuristic, the LLM identifies which column
    contains the primary entity identifier.
  \item \textbf{Relation extraction:}  For non-tabular content (prose
    paragraphs, nested lists), the LLM extracts
    (entity, relation, value) triples that the regex pipeline would miss.
\end{itemize}

\noindent
Critically, the hybrid mode preserves the core invariant: all numeric
values are stored with SHA-256 hashes and verified on lookup.  The LLM
assists with \emph{structural} decisions (which column is the entity?
what relation does this paragraph express?) but the \emph{values}
remain deterministically extracted and integrity-checked.  This
design ensures that even in hybrid mode, the ground truth is
verifiable against the source document.  Users select the ingestion
mode via the CLI (\texttt{-{}-ingest-mode default|hybrid}) or the
Python API (\texttt{ingest\_markdown(path, mode="hybrid")}).

\paragraph{Benchmark as a Development Tool.}
FinStructBench can serve as an evaluation harness during the development
of retrieval-augmented generation (RAG) systems, agentic frameworks and
tool-use architectures for financial documents.  The provably correct
ground truth eliminates the need for human evaluation, enabling rapid
iteration on system design.

\paragraph{Extensibility.}
The benchmark toolkit is designed for extensibility.  New question
categories can be added by implementing a generator class with a
\texttt{generate(graph)} method.  New instances can be created from
any markdown document.  The phase encoder system can be extended with
domain-specific threshold rules.

\paragraph{Limitations.}
FinStructBench currently uses synthetic financial documents rather
than real regulatory filings.  While the synthetic reports are
designed to be structurally realistic, real documents may contain
additional complexity (e.g., footnotes, cross-references to external
regulations, redacted sections).  We plan to extend the benchmark
with anonymized real-world instances in future work.

The benchmark evaluates a single LLM (Claude Sonnet~4) in a
zero-shot, full-context setting.  While we compare deterministic and
hybrid question modes, both use the same underlying ground truth.
Future work should evaluate multiple models, few-shot prompting,
chain-of-thought reasoning and tool-augmented architectures.

% =============================================================================
\section{Released Artifacts}
\label{sec:artifacts}

FinStructBench is released as an open-source Python package with
the following components:

\begin{enumerate}[leftmargin=2em]
  \item \textbf{Toolkit} (\texttt{finstructbench/}): Document ingester,
    ten question generators, scoring module, LLM caller and benchmark
    orchestrator.
  \item \textbf{Instances} (\texttt{instances/}): Five markdown financial
    reports with reproducible generator scripts.
  \item \textbf{Results} (\texttt{results/}): Baseline results with
    per-question details in JSON format.
  \item \textbf{CLI}: \texttt{python -m finstructbench run <doc.md>}
    with options for category limit, model and graph-only mode.
\end{enumerate}

% =============================================================================
\section{Conclusion}
\label{sec:conclusion}

We presented FinStructBench, a benchmark for structured information
retrieval from financial documents with provably correct ground truth.
Three sets of experiments yield complementary findings:

\begin{enumerate}[leftmargin=2em]
  \item \textbf{LLM evaluation.}  Across 454~questions spanning ten
    categories and five financial document types, Claude Sonnet~4
    achieves 58--59\% overall against the 100\% graph baseline, with a
    clear complexity gradient: 94\% on absence detection but only 30\%
    on contradiction and 32\% on counting.
  \item \textbf{Question mode comparison.}  Comparing deterministic
    (template-based) and hybrid (LLM-paraphrased) question modes reveals
    a marginal +1~pp improvement overall, with cross-reference queries
    benefiting most (+10~pp).
  \item \textbf{Ingestion mode comparison.}  A three-way comparison of
    regex-only, LLM-only and hybrid ingestion using Claude Opus~4
    demonstrates that LLM-only extraction recovers only 12.5\% of
    structured data (Jaccard~$= 0.015$), while hybrid ingestion preserves
    100\% of regex values exactly and adds 22\% more entries from prose
    extraction.
\end{enumerate}

\noindent
Together these results demonstrate that current LLMs are insufficient
for reliable financial document analysis: they struggle with structured
retrieval tasks (58\% accuracy) and cannot replace deterministic
extraction pipelines (12.5\% recovery).  Hybrid architectures that
combine rule-based precision with LLM flexibility offer the most
promising path forward.  FinStructBench provides a rigorous, extensible
evaluation framework for measuring progress toward this goal.

% =============================================================================
\bibliographystyle{plainnat}
\begin{thebibliography}{40}

\bibitem[Alvarado et~al.(2015)]{alvarado2015domain}
Alvarado, J.~C.~S., Verspoor, K., and Baldwin, T. (2015).
\newblock Domain adaption of named entity recognition to support credit risk
  assessment.
\newblock In \emph{Proceedings of the Australasian Language Technology
  Association Workshop}, pp.~84--90.

\bibitem[Edge et~al.(2024)]{edge2024graphrag}
Edge, D., Trinh, H., Cheng, N., Bradley, J., Chao, A., Mody, A.,
  Truitt, S., and Larson, J. (2024).
\newblock From local to global: A graph {RAG} approach to query-focused
  summarization.
\newblock \emph{arXiv preprint arXiv:2404.16130}.

\bibitem[Chen et~al.(2020)]{chen2020hybridqa}
Chen, W., Zha, H., Chen, Z., Xiong, W., Wang, H., and Wang, W.~Y. (2020).
\newblock {HybridQA}: A dataset of multi-hop question answering over tabular
  and textual data.
\newblock In \emph{Findings of EMNLP}, pp.~1026--1036.

\bibitem[Chen et~al.(2021)]{chen2021finqa}
Chen, Z., Chen, W., Smiley, C., Shah, S., Borber, I., Leu, S., Zhu, Y.,
  Kumar, S., Zhang, J., and Wang, W.~Y. (2021).
\newblock {FinQA}: A dataset of numerical reasoning over financial data.
\newblock In \emph{Proceedings of EMNLP}, pp.~7862--7872.

\bibitem[Chen et~al.(2022)]{chen2022convfinqa}
Chen, Z., Li, S., Smiley, C., Ma, Z., Shah, S., and Wang, W.~Y. (2022).
\newblock {ConvFinQA}: Exploring the chain of numerical reasoning in
  conversational finance question answering.
\newblock In \emph{Proceedings of EMNLP}, pp.~6279--6292.

\bibitem[Herzig et~al.(2020)]{herzig2020tapas}
Herzig, J., Nowak, P.~K., M\"uller, T., Piccinno, F., and
  Eisenschlos, J. (2020).
\newblock {TAPAS}: Weakly supervised table parsing via pre-training.
\newblock In \emph{Proceedings of ACL}, pp.~4320--4333.

\bibitem[Huang et~al.(2022)]{huang2022layoutlmv3}
Huang, Y., Lv, T., Cui, L., Lu, Y., and Wei, F. (2022).
\newblock {LayoutLMv3}: Pre-training for document {AI} with unified text and
  image masking.
\newblock In \emph{Proceedings of the 30th ACM International Conference on
  Multimedia}.

\bibitem[Iyyer et~al.(2017)]{iyyer2017search}
Iyyer, M., Yih, W., and Chang, M. (2017).
\newblock Search-based neural structured learning for sequential question
  answering.
\newblock In \emph{Proceedings of ACL}, pp.~1821--1831.

\bibitem[Jiang et~al.(2023)]{jiang2023structgpt}
Jiang, J., Zhou, K., Dong, Z., Ye, K., Zhao, X., and Wen, J.-R. (2023).
\newblock {StructGPT}: A general framework for large language model to reason
  over structured data.
\newblock In \emph{Proceedings of EMNLP}, pp.~9237--9251.

\bibitem[Kim et~al.(2022)]{kim2022donut}
Kim, G., Hong, T., Yim, M., Nam, J., Park, J., Yim, J., Hwang, W.,
  Yun, S., Han, D., and Park, S. (2022).
\newblock {OCR}-free document understanding transformer.
\newblock In \emph{European Conference on Computer Vision (ECCV)}.

\bibitem[Lee et~al.(2023)]{lee2023pix2struct}
Lee, K., Joshi, M., Turc, I., Hu, H., Liu, F., Eisenschlos, J.,
  Khandelwal, U., Shaw, P., Chang, M.-W., and Toutanova, K. (2023).
\newblock {Pix2Struct}: Screenshot parsing as pretraining for visual language
  understanding.
\newblock In \emph{Proceedings of ICML}, pp.~18893--18912.

\bibitem[Li et~al.(2024)]{li2024long}
Li, Y., Packer, C., and Gonzalez, J.~E. (2024).
\newblock How long can open-source {LLMs} truly promise on context length?
\newblock \emph{arXiv preprint arXiv:2402.17688}.

\bibitem[Liu et~al.(2022)]{liu2022tapex}
Liu, Q., Chen, B., Guo, J., Ziyadi, M., Lin, Z., Chen, W., and
  Lou, J.-G. (2022).
\newblock {TAPEX}: Table pre-training via learning a neural {SQL} executor.
\newblock In \emph{International Conference on Learning Representations (ICLR)}.

\bibitem[Malo et~al.(2014)]{malo2014good}
Malo, P., Sinha, A., Korhonen, P., Wallenius, J., and Takala, P. (2014).
\newblock Good debt or bad debt: Detecting semantic orientations in economic
  texts.
\newblock \emph{Journal of the Association for Information Science and
  Technology}, 65(4):782--796.

\bibitem[Mathew et~al.(2021)]{mathew2021docvqa}
Mathew, M., Karatzas, D., and Jawahar, C.~V. (2021).
\newblock {DocVQA}: A dataset for {VQA} on document images.
\newblock In \emph{Proceedings of WACV}, pp.~2200--2209.

\bibitem[Schick et~al.(2023)]{schick2023toolformer}
Schick, T., Dwivedi-Yu, J., Dess\`i, R., Raileanu, R., Lomeli, M.,
  Zettlemoyer, L., Cancedda, N., and Scialom, T. (2023).
\newblock Toolformer: Language models can teach themselves to use tools.
\newblock In \emph{Advances in Neural Information Processing Systems (NeurIPS)}.

\bibitem[Pasupat \& Liang(2015)]{pasupat2015compositional}
Pasupat, P. and Liang, P. (2015).
\newblock Compositional semantic parsing on semi-structured tables.
\newblock In \emph{Proceedings of ACL}, pp.~1470--1480.

\bibitem[Wang et~al.(2024)]{wang2024chainoftable}
Wang, Z., Zhang, H., Li, C.-L., Eisenschlos, J.~M., Perot, V., Wang, Z.,
  Miculicich, L., Fujii, Y., Shang, J., Lee, C.-Y., and Pfister, T. (2024).
\newblock Chain-of-Table: Evolving tables in the reasoning chain for table
  understanding.
\newblock In \emph{International Conference on Learning Representations (ICLR)}.

\bibitem[Yao et~al.(2023)]{yao2023react}
Yao, S., Zhao, J., Yu, D., Du, N., Shafran, I., Narasimhan, K., and
  Cao, Y. (2023).
\newblock {ReAct}: Synergizing reasoning and acting in language models.
\newblock In \emph{International Conference on Learning Representations (ICLR)}.

\bibitem[Ye et~al.(2023)]{ye2023dater}
Ye, Y., Hui, B., Yang, M., Li, B., Huang, F., and Li, Y. (2023).
\newblock Large language models are versatile decomposers: Decompose evidence
  and questions for table-based reasoning.
\newblock In \emph{Proceedings of the 46th International ACM SIGIR Conference
  on Research and Development in Information Retrieval}, pp.~174--184.

\bibitem[Shah et~al.(2022)]{shah2022flue}
Shah, R., Chawla, K., Eidnani, D., Shah, A., Du, W., Chava, S., Raman, N.,
  Smiley, C., Chen, J., and Yang, D. (2022).
\newblock When {FLUE} meets {FLANG}: Benchmarks and large pre-trained language
  model for financial domain.
\newblock In \emph{Proceedings of EMNLP}, pp.~2322--2335.

\bibitem[Zhang et~al.(2022)]{zhang2022subgraph}
Zhang, J., Chen, B., Zhang, L., Ke, X., and Ding, H. (2022).
\newblock Neural, symbolic and neural-symbolic reasoning on knowledge graphs.
\newblock \emph{AI Open}, 3:109--124.

\bibitem[Zhu et~al.(2021)]{zhu2021tat}
Zhu, F., Lei, W., Huang, Y., Wang, C., Zhang, S., Lv, J., Feng, F., and
  Chua, T. (2021).
\newblock {TAT-QA}: A question answering benchmark on a hybrid of tabular and
  textual content in finance.
\newblock In \emph{Proceedings of ACL}, pp.~3277--3287.

\bibitem[Butt et~al.(2024)]{butt2024benchagents}
Butt, N., Chandrasekaran, V., Joshi, N., Nushi, B., and
  Balachandran, V. (2024).
\newblock {BenchAgents}: Automated benchmark creation with multi-agent systems.
\newblock \emph{arXiv preprint arXiv:2410.22584}.

\bibitem[Dammu et~al.(2025)]{dammu2025dynamickgqa}
Dammu, P.~P.~S., Naidu, H., and Shah, C. (2025).
\newblock {Dynamic-KGQA}: A scalable framework for generating adaptive question
  answering datasets from knowledge graphs.
\newblock In \emph{Proceedings of the 48th International ACM SIGIR Conference
  on Research and Development in Information Retrieval}.

\bibitem[Gu et~al.(2021)]{gu2021grailqa}
Gu, Y., Kase, S., Vanni, M., Sadler, B., Liang, P., Yan, X., and
  Su, Y. (2021).
\newblock Beyond {I.I.D.}: Three levels of generalization for question
  answering on knowledge bases.
\newblock In \emph{Proceedings of The Web Conference (WWW)}, pp.~3477--3488.

\bibitem[Islam et~al.(2023)]{islam2023financebench}
Islam, P., Kannappan, A., Kiela, D., Qian, R., Scherrer, N., and
  Vidgen, B. (2023).
\newblock {FinanceBench}: A new benchmark for financial question answering.
\newblock \emph{arXiv preprint arXiv:2311.11944}.

\bibitem[Matlin et~al.(2026)]{matlin2026finforge}
Matlin, G., Theerthala, A., Gupta, A., JM, A., Castilla, R., Ng, Y.~M.,
  and Chava, S. (2026).
\newblock {FinForge}: Semi-synthetic financial benchmark generation.
\newblock \emph{arXiv preprint arXiv:2601.06747}.

\bibitem[Shashidhar et~al.(2025)]{shashidhar2025yourbench}
Shashidhar, S., Fourrier, C., Lozovskia, A., Wolf, T., Tur, G., and
  Hakkani-Tur, D. (2025).
\newblock {YourBench}: Easy custom evaluation sets for everyone.
\newblock In \emph{Proceedings of the Conference on Language Modeling (COLM)}.

\bibitem[Tang et~al.(2025)]{tang2025grapharena}
Tang, J., Zhang, Q., Li, Y., Chen, N., and Li, J. (2025).
\newblock {GraphArena}: Evaluating and exploring large language models on graph
  computation.
\newblock In \emph{International Conference on Learning Representations (ICLR)}.

\bibitem[Tatarinov et~al.(2025)]{tatarinov2025kgqagen}
Tatarinov, A., Kannan, A., Srinivasa, J., Raj, B., Anand, A., Singh, A.,
  Luthra, N., Lade, P., Shah, R., and Chava, S. (2025).
\newblock {KG-QAGen}: A knowledge-graph-based framework for systematic question
  generation and long-context {LLM} evaluation.
\newblock \emph{arXiv preprint arXiv:2505.12495}.

\bibitem[Trivedi et~al.(2017)]{trivedi2017lcquad}
Trivedi, P., Maheshwari, G., Dubey, M., and Lehmann, J. (2017).
\newblock {LC-QuAD}: A corpus for complex question answering over knowledge
  graphs.
\newblock In \emph{International Semantic Web Conference (ISWC)},
  LNCS vol.~10588, pp.~210--218. Springer.

\bibitem[Wu et~al.(2025)]{wu2025tablebench}
Wu, Y., Tang, J., and others (2025).
\newblock {TableBench}: A comprehensive and complex benchmark for table
  question answering.
\newblock In \emph{Proceedings of AAAI}, vol.~39, pp.~25497--25506.

\bibitem[Zhu et~al.(2025)]{zhu2025tableeval}
Zhu, J., Wang, J., Yu, B., Wu, X., Li, J., Wang, L., and Xu, N. (2025).
\newblock {TableEval}: A real-world benchmark for complex, multilingual, and
  multi-structured table question answering.
\newblock In \emph{Proceedings of EMNLP}.

\end{thebibliography}

\end{document}
